Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-08-31 paint-contour-tree.tex
Many changes in Section 5.2

\pagebreak
\newcommand{\pathTree}{P_{\mathcal{T}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{T}}}
\section{Lower Bound}

@@

\begin{observation}
 The contour tree of an minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
 to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
 factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
 manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
\end{observation}

@@

To do so, we now introduce the concept of path decompositions and their cost.  Then in the following two subsections 
we then relate path decompositions costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.

@@

Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$ corresponding to 
the path decomposition $P(T)$, and denote it by $\pathTree(T)$.

We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the path tree.  Similarly, for 
two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the path tree.

@@

We say two path trees $T_1$ and $T_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
between the two trees such that for any node $v\in T_1$ corresponding to a path of size $|p_v|$, $f(v)$ also corresponds to a path of size $|p_v|$.
For the purposes of the next lemma, we say two 2-manifolds are different, if their corresponding contour trees are the different.

@@

\begin{lemma}
\label{lem:pathDecomp}
Let $T$ be a merge tree, $P = P(T)$ any path decomposition of $T$, and $\pathTree$ the corresponding path tree.  
Then there are at least $\Pi_{p_i\in P} |p_i|!$ different 2-manifolds with an equivalent path tree.
\end{lemma}
\begin{proof}
 We describe a simple input mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
 this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
 whose size is that described in the lemma statement.

@@

 Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
 diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
 diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
 (since all these saddles have a direct line of sight to this apex).  In particular, 
 this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding path 
 tree is equivalent to $\pathTree$.

@@

\begin{corollary}
 Let $A(S)$ be any algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
 Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
 merge tree of $S$.
\end{corollary}
\begin{proof}
 Given an input surface $S$ whose merge tree has a particular path decomposition $P$, by the above lemma, there are $\Pi_{p_i\in P} |p_i|!$ other surfaces 
 with distinct merge trees, all of which have a path tree equivalent to $\pathTree$.  Therefore, any algorithm which computes the merge tree must distinguish between 
 these possibilities, and therefore its corresponding decision tree must have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, 
 this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
\end{proof}

@@

@@

Recall we are ultimately trying to show $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ for some path decomposition $P=P(T)$, 
or more specifically $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$.  In particular, if we could 
show for each path $p\in P$ that $\sum_{v\in p} \log |H_v|= |p|\log|p|$ then we would be done.
In other words we can think of each path as having a budget of $|p|\log|p|$, and ideally the budget 
of each path will pay for the algorithm's cost due to this path.  This is too good to hope for, so instead 
we will divide paths into two categories, called heavy and light, which we show below in \Lem{pathBounds} roughly corresponds 
to whether a path can pay for itself or not.

@@

We are now ready to define the maximum path decomposition, which will allow us to recharge the light paths to heavy paths.

@@

\begin{lemma}
\label{lem:geometric}
 Let $p$ and $q$ be two light paths such that $p$ is the parent of $q$ in the path tree $\pathTreeA$.  
 Then $|H_q|\leq \frac{3}{5} |H_p|$, where $H_p$ and $H_q$ are the heaps at the roots of $p$ and $q$, respectively. 
\end{lemma}
\begin{proof}
 Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
 i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
 As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
 the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
 Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
 Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
 Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
\end{proof}

@@

\begin{definition}
 Let $p$ be a path in a path decomposition $P(T)$, whose corresponding path tree is $\pathTree$.
 Call a light path $q\in P(T)$ light reachable from $p$ if the path from $q$ to $p$ in $\pathTree$ 
 contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
 of all light reachable paths from $p$ the \emph{light path tree} of $p$, and denote is $\mathcal{L}(p)$.
\end{definition}

In the following subsection we show how to exploit this geometric series behavior in light path trees.

\subsubsection{An asymptotically equivalent weight function}
Let $\mathcal{L} = \mathcal{L}(r)$ be a light path tree of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
The intuitive idea is that due to the geometric decreasing behavior of light path trees described in the previous subsection, 
one can show that for the entire light path tree our algorithm pays roughly the cost due to the root path $r$.  This argument 
is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.

For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $\mathcal{L}$.

@@

First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
was not already charged to descendant light paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
To do so, we first we augment $\mathcal{L}$ by attaching one 
leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  
Now we can define a weight function $w(u)$ over any node $u$ in the tree, which is defined to be the total cost of all leaves in its subtree.   
To simplify notation, we now use $\mathcal{L}$ to refer to this new weighted tree (rather than its unweighted counterpart).

As the weight $w(v)$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $v$, we have the following.

\begin{claim}
For any $v\in \mathcal{L}$, $w(v)\geq |H_v|$.
\end{claim}

@@

We now would like to show conversely that $w(v)= O(|H_v|)$.  
To do so we first introduce some notation.  
Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
that are contained in the subtree of $\mathcal{L}$ which is rooted at $v$.

First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
This holds by the same argument as in \Lem{adjacent}, namely, $w(u)$ by definition 
counts all the contributing heavy path heaps and initial leaf heaps that contribute 
to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
of vertices that could have been removed from one of these heaps before reaching the 
root of $v$ (since a vertex in a heap can only get removed if it appears on one of 
the paths).

@@

\begin{lemma}
Let $\mathcal{L}$ be the light path tree of some path $r$.  
Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
\end{lemma}
% \begin{proof}
% We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
% we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ where $p$ is the parent 
% of $v$ in $\mathcal{L}$.
% 
% Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
% (i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
% two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
% we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{distribution}, $|H_v|\geq \frac{4}{5} w(v)$.
% Combining this fact with \Lem{geometric} gives,
% \[
% w(v) \leq \frac{5}{4}|H_v| \leq \frac{5}{4} \frac{3}{5} |H_p| \leq \frac{3}{4} w(p).
% \]
% 
% \end{proof}
%
\begin{proof}
By the above claim, we know that $|H_v|\geq \frac{4}{5} w(v)$ if it holds that 
$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
So let $v$ be the lowest path in $\mathcal{L}$ such that this fractional decrease fails to hold, i.e. $w(c)> (4/5) w(v)$ 
where $c$ is a child of $v$ in $\mathcal{L}$, and conversely all vertices in $\mathcal{L}_c$ 
the fractional decrease holds.  As the fractional decrease holds for a paths in $\mathcal{L}_c$, 
we have $|H_c|\geq \frac{4}{5} w(c)$.  Therefore,
\[
w(c) \leq \frac{5}{4}|H_c| \leq \frac{5}{4} \frac{3}{5} |H_v| \leq \frac{3}{4} w(v).
\]
However, this is a contradiction with the assumption $w(c)> (4/5) w(v)$ , and hence the 
fractional decrease always holds implying the claim always holds.
\end{proof}

\begin{corollary}
 Let $\mathcal{L}$ be a light path tree of a light path $r$.  Then for any $v\in \mathcal{L}$, $w(v) = O(|H_v|)$.
\end{corollary}

%
%
% So $w(r)$ upper bound $H_r$, but by how much?  Consider a path $v\in \mathcal{L}$.  
% As we move from the leaf to the root of $v$ we now want to understand 
% how the heap can decreases.  Now the only way a vertex in the heap can leave that 
% heap is if that vertex is one of the vertices we traverse in $v$, i.e. the total 
% loss in vertices from the heap along $v$ is bounded by $|v|$.  
% Therefore, $|H_r|\geq W-\sum_{v\in \mathcal{L}} |v|$.  
% (This is an extension of \Lem{adjacent}, and probably should be merged with it.)
% As $\mathcal{L}$ consists of only light paths we have $|v|\leq \sqrt{|H_v|}/100 \leq \sqrt{w(v)}/100$, 
% and therefore $|H_r|\geq W-\sum_{v\in \mathcal{L}} \sqrt{w(v)}/100$.

\pagebreak
\subsection{Junk, don't read}

\begin{lemma}
 Let $\chi(T)$ be a valid coloring of a tree $T$, and let $\mathcal{P}(T)$ be the corresponding maximum path decomposition.
 Let $p$ be a light path in $\mathcal{P}(T)$ and let $LPT(p)$ be the light path tree of $p$.

 Then $\sum_{q\in L(p)} \sum_{v\in q} \log |H_v| = |H_p|\log |H_p|$.
\end{lemma}
\begin{proof}UNDER CONSTRUCTION / NOTES:

 Claim: Let $q$ be some path of depth $i$ in $LPT(p)$, and let $H_q$ be the heap at the root of this path.
 Then $|H_q|< |H_p|/c^i$ for some constant $c>1$.  This should follow from using a maximum path decomposition.

 This claim would imply the depth of the tree is $O(\log |H_p|)$ and the total number of paths in the tree is $O(|H_p|)$, 
 thought I am not sure if either of these statements are useful.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{More Junk}

The following can be viewed as stronger version of \Lem{adjacent}, which is possible as the path is now assumed to be maximum.

\begin{lemma}
 Let $\chi(T)$ be a valid coloring of a tree $T$, and let $\mathcal{P}(T)$ be the corresponding maximum path decomposition.
 Let $p$ be a light path in $\mathcal{P}(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.

 For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
  Then $\sum_{i=1}^k |H_i| \leq |H_p|/c$ for some constant $c>1$, where $H_p = H_{r_p}$.
\end{lemma}

% \begin{proof}UNDER CONSTRUCTION: 
% For simplicity, let $H_{k+1}$ denote the heap at the leaf vertex of $p$.  As the $H_k+1$ is disjoing from all other $H_{i}$, \Lem{adjacent} implies that 
% $H_p > \sum_{j=1}^{k+1} H_i - |p|$.  
% 
% Let $x$ be the largest index such that $H_x > H_p / 3$.
% By the above observation and since $p$ is a light path we have 
% \[
%  H_p > \sum_{j=1}^{k} H_i + H_{k+1} - |p|
% \] 
% \end{proof}

>>>>

\pagebreak
\newcommand{\pathTree}{P_{\mathcal{S}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
\section{Lower Bound}

@@

\begin{observation}
 The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
 to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
 factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
 manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
\end{observation}

@@

To do so, we now introduce the concept of path decompositions and their cost.  Then in the following two subsections 
we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.

@@

Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
the path decomposition $P(T)$, and denote it by $\pathTree(T)$.

We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub.

@@

We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.

@@

\begin{lemma}
\label{lem:pathDecomp}
Let $T$ be a merge tree, $P = P(T)$ any path decomposition of $T$, and $\pathTree$ the corresponding shrub.  
Then there are at least $\Pi_{p_i\in P} |p_i|!$ different 2-manifolds with an equivalent shrub.
\end{lemma}
\begin{proof}
 We describe a simple input mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
 this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
 whose size is that described in the lemma statement.

@@

 Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
 diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
 diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
 (since all these saddles have a direct line of sight to this apex).  In particular, 
 this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.

@@

\begin{corollary}
 Let $A(S)$ be any algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
 Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
 merge tree of $S$.
\end{corollary}
\begin{proof}
 Given an input surface $S$ whose merge tree has a particular path decomposition $P$, by the above lemma, there are $\Pi_{p_i\in P} |p_i|!$ other surfaces 
 with distinct merge trees, all of which have a shrub equivalent to $\pathTree$.  Therefore, any algorithm which computes the merge tree must distinguish between 
 these possibilities, and therefore its corresponding decision tree must have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, 
 this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
\end{proof}

@@

Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e. $H_p = H_{r_p}$.

@@

Recall we are ultimately trying to show $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ for some path decomposition $P=P(T)$, 
or more specifically $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$.  In particular, if we could 
show for each path $p\in P$ that $\sum_{v\in p} \log |H_v|= |p|\log|p|$ then we would be done.
In other words we can think of each path as having a budget of $|p|\log|p|$, and ideally the budget 
of each path will pay for the algorithm's cost due to this path.  This is too good to hope for, so instead 
we will divide paths into two categories, called heavy and light, which as we show below in \Lem{pathBounds} roughly corresponds 
to whether a path can pay for itself or not.

\begin{definition}
 Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
 For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{p}|}/100$, and we say it is \emph{heavy} otherwise.
\end{definition}

\begin{observation}
\label{obs:decrease}
 Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
 $w$ to the root has one less vertex than the path from $v$ to the root.
 In particular, we have the following more general property.
 Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $v$ is a descendant of $u$.  
 Then $|H_v| \leq |H_u| + d(u,v)$. 
\end{observation}

The following can be thought of as a generalization of the above observation, and will be useful in later sections.

\begin{lemma}
\label{lem:adjacent}
 Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
 Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$ in $\pathTree(T)$.
 For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
 Also, let $H_0$ denote the heap at the leaf of $p$.

@@

\begin{lemma}
\label{lem:pathBounds}
 Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
 Then, 
 \begin{compactenum}
  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
 \end{compactenum}
\end{lemma}
\begin{proof}
 By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
 As such, 
 \[
 \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|).
 \]
 If $p$ is a heavy path then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
 $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$.
\end{proof}

For a light path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
Our eventual goal will be to show that by choosing the right path decomposition, this excess 
can be recharged to heavy paths. In the next couple sections we introduce the appropriate path decomposition and prove 
some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.

@@

\begin{lemma}
\label{lem:geometric}
 Let $p$ and $q$ be two light paths such that $p$ is the parent of $q$ in the shrub $\pathTreeA$.  
 Then $|H_q|\leq \frac{3}{5} |H_p|$, where $H_p$ and $H_q$ are the heaps at the roots of $p$ and $q$, respectively. 
\end{lemma}
\begin{proof}
 Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
 i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
 As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
 the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
 Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
 Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
 Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
\end{proof}

@@

\begin{definition}
 Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
 Call a light path $q\in P(T)$ light reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
 contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
 of all light reachable paths from $p$ the \emph{light shrub} of $p$, and denote it $\mathcal{L}(p)$.
\end{definition}

In the following subsection we show how to exploit this geometric series behavior of light shrubs.

\subsubsection{An asymptotically equivalent weight function}
\label{sec:weight}
Let $\mathcal{L} = \mathcal{L}(r)$ be a light shrub of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
The intuitive idea is that due to the geometric decreasing behavior of light shrubs described in the previous subsection, 
one can show that for the entire light shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.

For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $\pathTreeA(T)$.

@@

First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
was not already charged to descendant light paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
To do so, we first augment $\mathcal{L}$ by attaching one 
leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  
Now we can define a weight function $w(u)$ over any node $u$ in the tree, which is defined to be the total cost of all leaves in its subtree.   
To simplify notation, we now use $\mathcal{L}$ to refer to this new weighted tree (rather than its unweighted counterpart).

As the weight $w(p$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $p$, we have the following.

\begin{claim}
For any $p\in \mathcal{L}$, $w(p)\geq |H_p|$.  Moreover, for any $v\in p$, $w(p)\geq |H_v|$.
\end{claim}

@@

We now would like to show conversely that $w(v)= O(|H_v|)$.  
To do so we first introduce some notation.  
Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
that are in the subtree of $\mathcal{L}$ which is rooted at $v$.

First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
This holds by the same argument as in \Lem{adjacent}, namely, $w(v)$ by definition 
counts all the contributing heavy path heaps and initial leaf heaps that contribute 
to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
of vertices that could have been removed from one of these heaps before reaching the 
root of $v$ (since a vertex in a heap can only get removed if it appears on one of 
the paths).

@@

\begin{lemma}
Let $\mathcal{L}$ be the light shrub of some path $r$.  
Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
\end{lemma}
% \begin{proof}
% We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
% we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ where $p$ is the parent 
% of $v$ in $\mathcal{L}$.
% 
% Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
% (i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
% two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
% we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{distribution}, $|H_v|\geq \frac{4}{5} w(v)$.
% Combining this fact with \Lem{geometric} gives,
% \[
% w(v) \leq \frac{5}{4}|H_v| \leq \frac{5}{4} \frac{3}{5} |H_p| \leq \frac{3}{4} w(p).
% \]
% 
% \end{proof}
%
\begin{proof}
By the above claim, we know that $|H_v|\geq \frac{4}{5} w(v)$ if it holds that 
$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
So let $v$ be the lowest path in $\mathcal{L}$ such that this fractional decrease fails to hold, i.e. $w(c)> (4/5) w(v)$ 
where $c$ is a child of $v$ in $\mathcal{L}$, and conversely all vertices in $\mathcal{L}_c$ 
the fractional decrease holds.  As the fractional decrease holds for a paths in $\mathcal{L}_c$, 
we have $|H_c|\geq \frac{4}{5} w(c)$.  Therefore,
\[
w(c) \leq \frac{5}{4}|H_c| \leq \frac{5}{4} \frac{3}{5} |H_v| \leq \frac{3}{4} w(v).
\]
However, this is a contradiction with the assumption $w(c)> (4/5) w(v)$ , and hence the 
fractional decrease always holds implying the claim always holds.
\end{proof}

\begin{corollary}
\label{cor:equivalent}
 Let $\mathcal{L}$ be a light shrub of a light path $r$.  Then for any $v\in \mathcal{L}$, $|H_v| \leq w(v) \leq \frac{5}{4} |H_v|$.
\end{corollary}

Now we are finally ready to make use of the properties we have proven about the weight function we introduced.

\Ben{The constant in the lemmas below, but it is no more than twice that.}

\begin{lemma}
\label{lem:lightCost}
 Let $\mathcal{L}$ be a light shrub of a light path $r$.  Let $V$ be the set of heavy paths from the full shrub which are adjacent to paths in $\mathcal{L}$.  
  Then $\sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q| \leq \frac{4}{3} |H_r|\log |H_r|$
\end{lemma}
\begin{proof}
\Ben{
The following is a proof for $\frac{3}{2}+\eps$, rather than $\frac{4}{3}$.  The discrepancy is due to a corrected calculation error.  
4/3 can be achieved by tightening constants above, and is left in the lemma statement as it is used later.
}

Recall that in a proper coloring of a binary tree $T$, the color of any vertex must be the color of one of its descendants.  
Therefore, any vertex in the light path shrub must appear in the heap of either the root of an adjacent 
path from the full shrub or from the heap of a leaf of a path in $\mathcal{L}$.  
Also note $|H_r|$ precisely counts sum of the number of vertices from these heaps that do not appear in $\mathcal{L}$.
As there are no adjacent light paths to a light shrub, the total number of vertices over all paths in $\mathcal{L}$ is at most 
$m = \sum_{p\in \mathcal{L}} |H_{p_f}| +\sum_{q\in V} |H_q| - |H_r| = w(r)-|H_r|$, where $H_{p_f}$ is the heap at the leaf of $p$.

Using this observation and \Cor{equivalent} we then have, 
\begin{align*}
 \sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q|
 &\leq \sum_{p\in \mathcal{L}}\sum_{v\in p} \log(w(r)) + \sum_{q\in V} |H_q|\log(w(r))\\
 &\leq \log(w(r))\pth{ w(r)-|H_r| } + \log(w(r))w(r)\\ 
 &= (2w(r)-|H_r|)\log(w(r)) \leq \pth{\frac{5}{2}|H_r|-|H_r|}\log\pth{\frac{5}{4}|H_r|}\\
 &= \frac{3}{2}|H_r|\log\pth{\frac{5}{4}|H_r|}
\end{align*}
\end{proof}

\subsubsection{Paying it down the line}

% \begin{claim}
% \label{clm:sizes}
%  Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
%  Let $H_f$ denote the heap at the leaf of $p$.  We then have the following
%  \begin{compactenum}
%   \item $|H_f|\leq (1+1/c)|H_p|$
%   \item $|H_f|\geq |H_p|/3$
%   \item For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
%  \end{compactenum} 
% \end{claim}
% \begin{proof}
%   For any vertex $v\in p$ \Obs{decrease} implies that $|H_v|\leq |H_p| + |p| < (1+1/c) |H_p|$.
%   First observe that by setting $v=f$ this immediately implies the first part of the claim.
%  % By a similar argument, $|H_f|\geq |H_p| - |p| > (1-1/c)|H_p|$.
%   
%   To prove the last part of the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
%   adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
%   $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
%   By \Obs{decrease} we then have, 
%    $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
%   When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
%   which is a contradiction as $|H_p|>c|p|>6$.
% \end{proof}

\Ben{I had originally thought the following claim would be useful in proving the lemma below.}
\begin{claim}
\label{clm:sizes}
 Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
 We then have the following
% \begin{compactenum}
%  \item 
  For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
%  \item If there does not exist and adjacent path $q$ such $H_q\geq H_p/3$, then $|H_f|\geq |H_p|/3$
% \end{compactenum} 
\end{claim}
\begin{proof}
  To prove the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
  adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
  $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
  By \Obs{decrease} we then have, 
   $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
  When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
  which is a contradiction as $|H_p|>c|p|>6$.  
\end{proof}
